---
title: "Final Project"
author: "Brian Butler "
date: "`r Sys.Date()`"
documentclass: article
geometry: margin=1in
fontsize: 11pt
output:
  pdf_document:
    toc: false
    df_print: kable
    fig_caption: false
    number_sections: false
    dev: pdf
    highlight: tango
  html_document:
    theme: default
    self_contained: true
    toc: false
    df_print: kable
    fig_caption: false
    number_sections: false
    smart: true
    dev: svg
---

```{r setup, include = FALSE}
# DO NOT ALTER THIS CHUNK
# Set knitr options
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  fig.width = 5,
  fig.asp = 0.618,
  out.width = "70%",
  dpi = 120, 
  fig.align = "center",
  cache = FALSE
)
# Load required packages
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(infer))
suppressPackageStartupMessages(library(modelr))
suppressPackageStartupMessages(library(broom))
# Load dataset
college <- read_rds("college.rds") %>%
  type_convert(
    na=combine("", "NA", "PrivacySuppressed")
  )
# Set seed
set.seed(98261936)
```

## Introduction

linear regression Test 

Is there a correlation between the number of branch and college acceptance rates?
in this analysis ?

In this analysis, I am going to explore the relationship between the number of
university branches (numbranch) and college acceptance rates (adm_rate). The
explanatory variable is the number of branches, which is a continuous variable 
and the response variable is the acceptance rate, also a continuous variable. To 
examine this relationship, I would carry out a linear regression test to 
determine whether there exists a statistically significant correlation between 
these two variables.

I expect that this comparison will help reveal how the spread of university 
branches can affect the admission prospects of students. While it is generally 
believed that the reputation and selectivity of a university have the most 
bearing on acceptance rates, analyzing the role of having multiple branches 
can provide a different perspective. Specifically, universities with more 
branches might provide students with more options for acceptance, which can 
change acceptance rates in general.


## Preprocessing


```{r}
college_data <- readRDS("college.rds")

college <- college_data %>%
  select(INSTNM, NUMBRANCH, ADM_RATE, REGION, CITY, STABBR, HIGHDEG, 
         SATVR25, SATMT25, ACTCM25)

college_reduced <- college %>%
  rename(
    institution = INSTNM,
    num_branch = NUMBRANCH,
    adm_rate = ADM_RATE,
    region = REGION,
  )

college_reduced <- college_reduced %>%
  mutate(
    region = recode(
      region,
      `0` = "U.S. Service Schools",
      `1` = "New England",
      `2` = "Mid East",
      `3` = "Great Lakes",
      `4` = "Plains",
      `5` = "Southeast",
      `6` = "Southwest",
      `7` = "Rocky Mountains",
      `8` = "Far West",
      `9` = "Outlying Areas"
    )
  )
```


## Visualization

I would like to examine what the relationship between the number of branches and 
admission rates, and I believe a a scatter plot does that adequately. 
```{r}
college_reduced %>%
  ggplot(aes(x = num_branch, y = adm_rate)) +
  geom_point(color = "blue", alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "green")
  labs(
    title = "Scatterplot of Number of Branches vs. Admission Rate",
    x = "Number of Branches", 
    y = "Admission Rate"
  ) +
  theme_minimal()
```

The scatter plot suggests that, on first impression, the number of branches 
does indeed rise with increased rates of admissions, albeit that this trend is 
not one hundred percent consistent as can be seen in the spread of blue dots. 
Nevertheless, beyond approximately 20 or so branches, the trend appears to level 
out and, in fact, to decline according to the green line. This is a hypothesis 
that the positive impact of increasing the number of branches on the admissions 
rate would have diminishing returns, and beyond a certain point, adding more 
branches might not lead to an increase, or could even be linked to a marginal 
decline in the admissions rate.


I am using this graph to to show the distribution across different institutions 
for admission rates. 
```{r}
college_reduced %>%
  ggplot(aes(x = adm_rate)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "green") +
  labs(
    title = "Distribution of Admission Rates",
    x = "Admission Rate",
    y = "Number of Institutions"
  ) +
  theme_minimal()
```

By looking at the histogram of the distribution of admission rates, we see that 
the distribution's peak is shifted in the direction of more admissions, 
with the mode being at 0.75, meaning most institutions admit about 
three-quarters of their applicants. It is unimodal and slightly right-skewed, 
meaning that there is a cluster of institutions with high admissions and a tail 
that shifts to the left in terms of lower admissions. One trend that is 
particularly notable is the decreasing number of institutions with admission 
rates falling below 0.75. There are very few institutions at the two extremes, 
namely highly selective institutions admitting less than 0.25 and admitting 
nearly all applicants (close to 1.00), which are likely to be possible outliers 
from the overall cluster. This distribution reflects a pattern in which a 
considerable number of institutions have fairly high acceptance rates, while the
truly highly selective institutions are less common among this sample.



I used a faceted scatter plot to see the relationship between branches and 
admissions rates to see how they differ by region, checking to see if it is a 
factor to consider. 
```{r}
college_reduced %>%
  ggplot(aes(x = num_branch, y = adm_rate)) + 
  geom_point(alpha = 0.3, color = "purple") +
  facet_wrap(~ region) +
  labs(
    title = "Admission Rate vs. Branches by the Region",
    x = "Branch Number ",
    y = "Admission Rate"
  ) +
  theme_minimal()
```
Referring to the faceted scatter plot, that breaks up the relationship between 
the number of branches and admission rate across regions, we can observe 
distinct patterns by regions. In each regional subplot, we can observe the range
of admission rates for a given number of branches. For instance, some regions, 
e.g., the 'Far West' and 'Great Lakes', have a wider spread of admission rates 
than others, e.g., 'New England' or the 'Southeast', whose admission rates are 
more clustered. Further, the regional central tendency of admission rates 
varies; some regions have institutions whose admission rates are higher on 
average than others. We can also see differing levels of the apparent 
correlation of the two variables. Whereas some regions show a weak positive or
negative trend, other regions show no recognizable linear relationship. 
The 'I.S. Service School' region is particularly sparse with very few data 
points, and it is difficult to see any recognizable patterns. By reporting the 
data in this way, it is possible to see quite easily that the relationship 
between branch number and admission rate is not the same everywhere, which 
 proves the influence of geographic context.


## Summary Statistics

This  would tell me the amount of colleges per region.
```{r}
college_reduced %>%
  group_by(region) %>%
  summarise(count = n ()) %>%
  arrange(desc(count))
```

```{r}
college_reduced %>%
  group_by(region) %>%
  summarise(
    count = n(),
    mean = mean(adm_rate, na.rm = TRUE),
    median = median(adm_rate, na.rm = TRUE),
    min  = min(adm_rate, na.rm = TRUE),
    max = max(adm_rate, na.rm = TRUE),
    sd = sd(adm_rate, na.rm = TRUE),
    iqr = IQR(adm_rate, na.rm = TRUE),
    range = max(adm_rate, na.rm = TRUE) - min(adm_rate, na.rm = TRUE)
  )
```

```{r}
college_reduced %>%
  summarise(
    count = n(),
    mean = mean(adm_rate, na.rm = TRUE),
    median = median(adm_rate, na.rm = TRUE),
    min  = min(adm_rate, na.rm = TRUE),
    max = max(adm_rate, na.rm = TRUE),
    sd = sd(adm_rate, na.rm = TRUE),
    iqr = IQR(adm_rate, na.rm = TRUE),
    range = max(adm_rate, na.rm = TRUE) - min(adm_rate, na.rm = TRUE)
  )
```
## Data Analysis

```{r}
model <- lm(adm_rate ~ num_branch, data = college_reduced)

tidy(model)
glance(model)
```


The model has a statistically significant though weak positive relationship 
between the number of branches and the outcome from the low p-value for 
'num_branch' (1.51e-06) and its positive coefficient (around 0.0072). This shows 
that the estimated outcome slightly increases as the number of branches 
increases. But the extremely low R-squared (about 0.01) in the above output 
tells us that this relationship explains only a very minor fraction of the total 
variation in the outcome. The intercept of about 0.67 is the predicted outcome 
when the number of branches is zero. That is, while the effect of the number of 
branches is statistically significant, its real effect on the prediction of the 
outcome from this model alone is extremely small.


```{r}
college_clean <- college_reduced %>%
  filter(!is.na(adm_rate), !is.na(num_branch))

college_clean %>%
  mutate(predicted = predict(model)) %>%
  ggplot(aes(x = predicted, y = adm_rate)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "pink") +
  labs(
    title = "Observed vs. Predicted Acceptance Rate",
       x = "Predicted",
       y = "Observed") +
  theme_minimal()
```
This graph compares the observed admission rates to the predicted rates based on 
our model. Each black dot is an institution, with its x-coordinate displaying 
the predicted rate and its y-coordinate displaying the actual observed rate. 
Ideally, if our model were ideal, all the points would lie along the pink line 
(where predicted = observed). But we can notice a lot of scatter of points 
around this line that informs us that our model's predictions are not quite 
right at the level of individual institutions. Even though there may be some 
sort of trend being pointed to by the spread of the points, the large spread 
would imply that other variables not outlined by our model influence admission 
rates.
 
```{r}
augment(model) %>%
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "pink") +
  labs(title = "Residuals vs. Fitted Values",
       x = "Fitted Values",
       y = "Residuals")
```
This graph plots the residuals on the y-axis against the fitted values from our 
model on the x-axis. If our model was a decent fit, we would hope that the 
residuals appeared randomly scattered about the horizontal red line at zero with 
no distinguishable patterns. We notice several things here, however. First, there 
is a bunching of points on the left, indicating more variability in the 
residuals for lower fitted values. Second, there may be some faint patterns or 
curvature in the spread of the residuals as the fitted values get larger, but 
it's not very pronounced. Ideally, the spread of the residuals should be 
approximately constant across all fitted values (homoscedasticity). The 
departures from a scatter around zero would indicate that possibly our model 
may not be detecting all the inherent patterns in data, and how accurate our 
prediction is may also depend on which admission rate is being predicted.

```{r}
qqnorm(residuals(model))
qqline(residuals(model), col = "purple")
```
This Q-Q plot indicates that our model's errors are not totally normally 
distributed. The clear curve of the points away from the straight line, 
specifically at the tails, indicates a departure from normality. Therefore, 
statistical tests assuming normally distributed errors are to be viewed with 
caution for this model.


## Conclusion

This analysis tested whether there is correlation between college acceptance 
rates and number of branches that an institution possesses. I observed high 
variation between institutions and possible weak positive correlation with 
number of branches by graphical inspection and summary statistics. This was 
supported by a linear model with a statistically significant but extremely small 
positive coefficient of num_branch. The low p-value indicates there is an 
effect, but the close to zero value of the measure of fit which was R-squared 
measure ~0.01 so it would indicate that number of branches explains virtually 
none of the variance of acceptance rates. What this means is that institutions 
with higher numbers of branches have slightly higher acceptance rates, but this 
effect of this variable itself is incredibly small.

Further model assumption testing showed that model predictive power was low. 
Observed vs. predicted identified widespread scattering, and residuals vs. 
fitted identified possible heteroscedasticity, i.e., that residuals of the model 
vary as fitted values vary. The Q-Q plot similarly identified that residuals 
were not normally distributed, especially at both ends of the tail, at the 
expense of validity of inference. All of these confirm that there exist other, 
and larger, variables involved in determining acceptance rates such as 
institution reputation, student population, and facilities. Essentially, college 
acceptance rates barely depend on college number of branches. This research 
contributes to overall awareness that college admissions depend on a complicated 
array of variables and just not institutional size.
